Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-09-06 Documents/M1/realizableFermionicM1/sections/imex.tex
Various edits.  Replaces homogeneous sphere figures

Eq.~\eqref{eq:ode} is to be integrated forward in time with an ODE integrator, and we begin with the following definition of a constraint-preserving time integration scheme, as applied to integrate a physical model forward in time.  
\begin{define}
  Given sufficient conditions, a constraint-preserving time integration scheme preserves the constraints of a model if the set of admissible states satisfying the constraints form a convex set.  
  \label{def:constraintPreserving}
\end{define}
As an example, given a spatial discretization of a model that is constraint-preserving with the forward Euler method for a sufficiently small time step, high-order explicit strong stability-preserving Runge-Kutta (SSP-RK) methods, which can be expressed as convex combinations of forward Euler steps, also preserve the constraints (possibly with a smaller time step), and form a class of constraint-preserving schemes.

@@

Unfortunately, the correction step in Eq.~\eqref{eq:imexCorrection} deteriorates the accuracy of the IMEX scheme when applied to the moment equations in the diffusion limit.  
The diffusion limit is characterized by frequent collisions ($\tau\ll 1$) in a purely scattering medium ($\xi=0$), and exhibits long-time behavior governed by (e.g., \cite{jinLevermore_1996})
\begin{equation}
  \pd{\cJ}{t} + \nabla\cdot\vect{\cH} = 0
  \quad\text{and}\quad
  \vect{\cH} = - \tau\,\nabla\cdot\vect{\cK}.  
  \label{eq:diffusionLimit}
\end{equation}
Here the time derivative term in the equation for the particle flux has been dropped (formally $\cO(\tau^{2})$) so that, to leading order in $\tau$, the second equation in Eq.~\eqref{eq:diffusionLimit} states a balance between the transport term and the collision term.  
Furthermore, in the diffusion limit, the distribution function is nearly isotropic so that $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$ and $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$.  
The absence of the transport operator in the correction step in Eq.~\eqref{eq:imexCorrection}, destroys the balance between the transport term and the collision term.  
We demonstrate the inferior performance of IMEX schemes with this correction step in the diffusion limit in Section~\ref{sec:smoothProblems}.  
We have also implemented and tested one of the IMEX schemes in Chertock et al. \cite{chertock_etal_2015} (not included in Section~\ref{sec:smoothProblems}), where the transport operator is part of the correction step, and found it to perform very well in the diffusion limit.  
However, we have not been able to prove the realizability-preserving property with this approach without invoking a too severe time step restriction.  
We therefore proceed to design constraint-preserving IMEX schemes, without the correction step, that perform better in the diffusion limit.  
We limit the scope to IMEX schemes of type~ARS.  
(It can be shown that IMEX schemes of type~A conforming to Definition~\ref{def:PD-IMEX} below do not exist; cf. \ref{app:noTypeA}.)

We take a heuristic approach to determine conditions on the coefficients of the IMEX scheme to ensure it performs well in the diffusion limit.  
Define the vectors 
\begin{equation}
  \vec{\cJ}=(\cJ^{(1)},\ldots,\cJ^{(s)})^{T}
  \quad\text{and}\quad
  \vec{\vect{\cH}}=(\vect{\cH}^{(1)},\ldots,\vect{\cH}^{(s)})^{T}.  
\end{equation}
We can then write the stages of the IMEX scheme in Eq.~\eqref{imexStages} applied to the number density equation as
\begin{equation}
  \vec{\cJ}=\cJ^{n}\,\vec{\vect{e}} - \dt\,\tilde{A}\,\nabla\cdot\vec{\vect{\cH}}, 
  \label{eq:numberDensityIMEX}
\end{equation}
where $\vec{\vect{e}}$ is a vector of length $s$ containing all ones, and the divergence operator acts individually on the components of $\vec{\vect{\cH}}$.  
Similarly, for the number flux equation we have
\begin{equation}
  \vec{\vect{\cH}}=\vect{\cH}^{n}\,\vec{\vect{e}}-\dt\,\big(\,\f{1}{3}\tilde{A}\,\nabla\vec{\cJ}+\f{1}{\tau}\,A\,\vec{\vect{\cH}}\,\big).
  \label{eq:numberFluxIMEX}
\end{equation}
In the context of IMEX schemes, the diffusion limit (cf. the second equation in \eqref{eq:diffusionLimit}) implies that the relation $A\,\vec{\vect{\cH}}=-\f{1}{3}\,\tau\,\tilde{A}\,\nabla\vec{\cJ}$ should hold.  
Define the pseudoinverse of the implicit coefficient matrix for IMEX schemes of type~ARS as
\begin{equation*}
    A^{-1}
    =\left(
        \begin{matrix}
          0 & 0 \\ 
          0 & \hat{A}^{-1}
        \end{matrix}
      \right).
\end{equation*}
For the stages $i=1,\ldots,s$ we then have
\begin{equation}
  \vect{\cH}^{(i)}=-\f{1}{3}\,\tau\,\vec{\vect{e}}_{i}^{T}A^{-1}\tilde{A}\,\vec{\vect{e}}\,\nabla\cJ^{n}+\cO(\dt\,\tau^{2}),
\end{equation}
where $\vec{\vect{e}}_{i}$ is the $i$th column of the $s\times s$ identity matrix, and we have introduced the expansion $\vec{\cJ}=\cJ^{n}\vec{\vect{e}}+\cO(\dt\,\tau)$.  
Thus, for $\vect{\cH}^{(i)}$ to be accurate in the diffusion limit, we require the IMEX coefficients to satisfy
\begin{equation}
  \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=2,\ldots,s.
  \label{eq:diffusionCondition}
\end{equation}
(The case $i=1$ is trivial and does not place any constraints on the IMEX coefficients.)  
Then, for the particle number equation we have
\begin{equation}
  \f{\cJ^{(s)}-\cJ^{n}}{\dt}
  =-\tilde{\vect{w}}^{T}(\nabla\cdot\vec{\vect{\cH}})
  =\f{1}{3}\tau\,\nabla^{2}\cJ^{n}+\cO(\dt\,\tau^{2}), 
\end{equation}
which approximates a diffusion equation for $\cJ$ with the correct diffusion coefficient $\tau/3$.  
(For a GSA IMEX scheme without the correction step, $\cJ^{n+1}=\cJ^{(s)}$.)

@@

Here we analyze the absolute stability of the proposed IMEX schemes, PA2+ and PD-ARS (given in \ref{app:butcherTables}), following \cite{hu_etal_2018}.  
As is commonly done, we do this in the context of the linear scalar equation
\begin{equation}
  \dot{u}=\lambda_{1}\,u+\lambda_{2}\,u,
  \label{eq:scalarODE}
\end{equation}
where $\lambda_{1}\in\mathbb{C}$ and real $\lambda_{2}\le0$.  
On the right-hand side of Eq.~\eqref{eq:scalarODE}, the first (oscillatory) term is treated explicitly, while the second (damping) term is treated implicitly.  
The IMEX schemes can then be written as $u^{n+1} =P(z_{1},z_{2})\,u^{n}$, where $P(z_{1},z_{2})$ is the amplification factor of the scheme, $z_{1}=\dt\,\lambda_{1} = x +iy$, and $z_{2} = \dt\,\lambda_{2} \le 0$.  
Stability of the IMEX scheme requires $|P(z_1,z_2)|\leq 1$.  
The stability regions of PA2+ and PD-ARS are plotted in Figure~\ref{fig:AbsoluteStability}.  
As can be seen from the figures, the absolute stability region of both schemes increases with increasing $|z_{2}|$ (increased damping).  
For a given $|z_{2}|\ge0$, the stability region of PD-ARS is larger than that of PA2+.  
In the linear case, a time step that satisfies the absolute stability for the explicit part of the IMEX scheme ($|z_2| = 0$) fulfills the stability requirement for the IMEX scheme as whole ($|z_2| \geq 0$).  
This holds for all PD-ARS schemes with $\epsilon\in [0,0.5)$.
\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth]{figures/AbsoluteStabilityPARSD}
    \includegraphics[width=0.5\textwidth]{figures/AbsoluteStabilityPA2+}
  \end{tabular}
   \caption{Boundary of the absolute stability region for different values of $z_2$ for PD-ARS with $\epsilon = 0.1$ (left panel) and PA2+ (right panel).}
  \label{fig:AbsoluteStability}
\end{figure}

>>>>

Eq.~\eqref{eq:ode} is to be integrated forward in time with an ODE integrator, and we begin with the following definition of a constraint-preserving time integration scheme, as applied to integrate a physical model forward in time.  
\begin{define}
  Given sufficient conditions, a constraint-preserving time integration scheme preserves the constraints of a model if the set of admissible states satisfying the constraints forms a convex set.  
  \label{def:constraintPreserving}
\end{define}
As an example, given a spatial discretization of a model that is constraint-preserving with the forward Euler method for a sufficiently small time step, high-order explicit strong stability-preserving Runge-Kutta (SSP-RK) methods, which can be expressed as convex combinations of forward Euler steps, also preserve the constraints (possibly with a smaller time step), and form a class of constraint-preserving schemes.

@@

Unfortunately, the correction step in Eq.~\eqref{eq:imexCorrection} deteriorates the accuracy of the IMEX scheme when applied to the moment equations in the diffusion limit.  
The diffusion limit is characterized by frequent collisions ($\tau\ll 1$) in a purely scattering medium ($\xi=0$), and exhibits long-time behavior governed by (e.g., \cite{jinLevermore_1996})
\begin{equation}
  \pd{\cJ}{t} + \nabla\cdot\vect{\cH} = 0
  \quad\text{and}\quad
  \vect{\cH} = - \tau\,\nabla\cdot\vect{\cK}.  
  \label{eq:diffusionLimit}
\end{equation}
Here the time derivative term in the equation for the particle flux has been dropped (formally $\cO(\tau^{2})$) so that, to leading order in $\tau$, the second equation in Eq.~\eqref{eq:diffusionLimit} states a balance between the transport term and the collision term.  
Furthermore, in the diffusion limit, the distribution function is nearly isotropic so that $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$ and $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$.  
The absence of the transport operator in the correction step in Eq.~\eqref{eq:imexCorrection}, destroys the balance between the transport term and the collision term.  
We demonstrate the inferior performance of IMEX schemes with this correction step in the diffusion limit in Section~\ref{sec:smoothProblems}.  
We have also implemented and tested one of the IMEX schemes in Chertock et al. \cite{chertock_etal_2015} (not included in Section~\ref{sec:smoothProblems}), where the transport operator is part of the correction step, and found it to perform very well in the diffusion limit.  
However, we have not been able to prove the realizability-preserving property with this approach without invoking a too severe time step restriction.  
We therefore proceed to design constraint-preserving IMEX schemes, without the correction step, that perform well in the diffusion limit.  
We limit the scope to IMEX schemes of type~ARS.  
(It can be shown that IMEX schemes of type~A conforming to Definition~\ref{def:PD-IMEX} below do not exist; cf. \ref{app:noTypeA}.)

We take a heuristic approach to determine conditions on the coefficients of the IMEX scheme to ensure it performs well in the diffusion limit.  
We leave the spatial discretization unspecified.  
Define the vectors 
\begin{equation}
  \vec{\cJ}=(\cJ^{(1)},\ldots,\cJ^{(s)})^{T}
  \quad\text{and}\quad
  \vec{\vect{\cH}}=(\vect{\cH}^{(1)},\ldots,\vect{\cH}^{(s)})^{T}.  
\end{equation}
(The components of $\vec{\cJ}$ and $\vec{\vect{\cH}}$ can, e.g., be the cell averages evolved with the DG method.)  
We can then write the stages of the IMEX scheme in Eq.~\eqref{imexStages} applied to the number density equation as
\begin{equation}
  \vec{\cJ}=\cJ^{n}\,\vec{\vect{e}} - \dt\,\tilde{A}\,\nabla\cdot\vec{\vect{\cH}}, 
  \label{eq:numberDensityIMEX}
\end{equation}
where $\vec{\vect{e}}$ is a vector of length $s$ containing all ones, and the divergence operator acts individually on the components of $\vec{\vect{\cH}}$.  
Similarly, for the number flux equation we have
\begin{equation}
  \vec{\vect{\cH}}=\vect{\cH}^{n}\,\vec{\vect{e}}-\dt\,\big(\,\f{1}{3}\tilde{A}\,\nabla\vec{\cJ}+\f{1}{\tau}\,A\,\vec{\vect{\cH}}\,\big).
  \label{eq:numberFluxIMEX}
\end{equation}
In the context of IMEX schemes, the diffusion limit (cf. the second equation in \eqref{eq:diffusionLimit}) implies that the relation $A\,\vec{\vect{\cH}}=-\f{1}{3}\,\tau\,\tilde{A}\,\nabla\vec{\cJ}$ should hold.  
Define the pseudoinverse of the implicit coefficient matrix for IMEX schemes of type~ARS as
\begin{equation*}
    A^{-1}
    =\left(
        \begin{matrix}
          0 & 0 \\ 
          0 & \hat{A}^{-1}
        \end{matrix}
      \right).
\end{equation*}
For the stages $i=1,\ldots,s$ we then have
\begin{equation}
  \vect{\cH}^{(i)}=-\f{1}{3}\,\tau\,\vec{\vect{e}}_{i}^{T}A^{-1}\tilde{A}\,\vec{\vect{e}}\,\nabla\cJ^{n}+\cO(\dt\,\tau^{2}),
\end{equation}
where $\vec{\vect{e}}_{i}$ is the $i$th column of the $s\times s$ identity matrix, and we have introduced the expansion $\vec{\cJ}=\cJ^{n}\vec{\vect{e}}+\cO(\dt\,\tau)$.  
Thus, for $\vect{\cH}^{(i)}$ to be accurate in the diffusion limit, we require the IMEX coefficients to satisfy
\begin{equation}
  \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=2,\ldots,s.
  \label{eq:diffusionCondition}
\end{equation}
(The case $i=1$ is trivial and does not place any constraints on the IMEX coefficients.)  
Then, for the particle number equation we have
\begin{equation}
  \f{\cJ^{(s)}-\cJ^{n}}{\dt}
  =-\tilde{\vect{w}}^{T}(\nabla\cdot\vec{\vect{\cH}})
  =\f{1}{3}\tau\,\nabla^{2}\cJ^{n}+\cO(\dt\,\tau^{2}), 
\end{equation}
which approximates a diffusion equation for $\cJ$ with the correct diffusion coefficient $\tau/3$.  
(For a GSA IMEX scheme without the correction step, $\cJ^{n+1}=\cJ^{(s)}$.)

@@

Here we analyze the absolute stability of the proposed IMEX schemes, PA2+ and PD-ARS (given in \ref{app:butcherTables}), following \cite{hu_etal_2018}.  
As is commonly done, we do this in the context of the linear scalar equation
\begin{equation}
  \dot{u}=\lambda_{1}\,u+\lambda_{2}\,u,
  \label{eq:scalarODE}
\end{equation}
where $\lambda_{1}\in\mathbb{C}$ and real $\lambda_{2}\le0$.  
On the right-hand side of Eq.~\eqref{eq:scalarODE}, the first (oscillatory) term is treated explicitly, while the second (damping) term is treated implicitly.  
The IMEX schemes can then be written as $u^{n+1} =P(z_{1},z_{2})\,u^{n}$, where $P(z_{1},z_{2})$ is the amplification factor of the scheme, $z_{1}=\dt\,\lambda_{1} = x +iy$, and $z_{2} = \dt\,\lambda_{2} \le 0$.  
Stability of the IMEX scheme requires $|P(z_1,z_2)|\leq 1$.  
The stability regions of PA2+ and PD-ARS are plotted in Figure~\ref{fig:AbsoluteStability}.  
(Contours of constant $z_{2}$ are included, and the stability region for a given $z_{2}$ is enclosed by the corresponding contour.)  
As can be seen from the figures, the absolute stability region of both schemes increases with increasing $|z_{2}|$ (increased damping).  
For a given $|z_{2}|>0$, the stability region of PD-ARS is larger than that of PA2+.  
In the linear case, a time step that satisfies the absolute stability for the explicit part of the IMEX scheme ($|z_2| = 0$) fulfills the stability requirement for the IMEX scheme as whole ($|z_2| \geq 0$).  
This holds for all PD-ARS schemes with $\epsilon\in [0,0.5)$.
\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth]{figures/AbsoluteStabilityPARSD}
    \includegraphics[width=0.5\textwidth]{figures/AbsoluteStabilityPA2+}
  \end{tabular}
   \caption{Boundary of the absolute stability region for different values of $z_2$ for PD-ARS with $\epsilon = 0.1$ (left panel) and PA2+ (right panel).}
  \label{fig:AbsoluteStability}
\end{figure}

### 2018-09-06 Documents/M1/realizableFermionicM1/sections/intro.tex
Various edits.  Replaces homogeneous sphere figures

In this paper we design numerical methods to solve the two-moment model governing transport of particles obeying by Fermi-Dirac statistics (e.g., neutrinos), with the ultimate target being nuclear astrophysics applications (e.g., neutrino transport in core-collapse supernovae and compact binary mergers).  
The numerical method is based on the discontinuous Galerkin (DG) method for spatial discretization, implicit-explicit (IMEX) methods for time integration, and is designed to preserve certain physical constraints of the underlying model.  
The latter property is achieved by considering the spatial and temporal discretization together with the closure procedure for the two-moment model.

In many applications, the particle mean free path is comparable to or exceeds other characteristic length scales in the system under consideration (i.e., the Knudsen number exceeds unity), and non-equilibrium effects may become important.  
In these situations, a kinetic description may be required, where particle distribution function $f(\vect{p},\vect{x},t)$ is solved for.  
The distribution function, a phase space density depending on momentum $\vect{p}\in\bbR^{3}$ and position $\vect{x}\in\bbR^{3}$, gives at time $t\in\bbR^{+}$ the number of particles in the phase space volume element $d\vect{p}\,d\vect{x}$ (i.e., $dN=f\,d\vect{p}\,d\vect{x}$).  
The evolution of the distribution function is governed by the Boltzmann equation, which states a balance between phase space advection and collisions (see, e.g., \cite{braginskii_1965,chapmanCowling_1970,lifshitzPitaevskii_1981}).

@@

Here we consider a two-moment model for particles governed by Fermi-Dirac statistics.  
It is well known from the two-moment model for particles governed by Maxwell-Boltzmann statistics (``classical'' particles with $f\ge0$), that the particle density is nonnegative and the magnitude of the flux vector is bounded by the particle density.  
(There are further constraints on the components of the Eddington tensor \cite{levermore_1984}.)  
Furthermore, the set of realizable moments generated by the particle density and flux vector constitutes a convex cone \cite{olbrant_etal_2012}.  
In the fermionic case, there is also an upper bound on the distribution function (e.g., $f\le1$) because Pauli's exclusion principle prevents particles from occupying the same microscopic state.  
The fermionic two-moment model has recently been studied theoretically in the context of maximum entropy closures \cite{lareckiBanach_2011,banachLarecki_2013,banachLarecki_2017b} and Kershaw-type closures \cite{banachLarecki_2017a}.  
Because of the upper bound on the distribution function, the algebraic constraints on realizable moments differ from the classical case with no upper bound, and can lead to significantly different dynamics when the occupancy is high (i.e., when $f$ is close to its upper bound).  
In the fermionic case, the set of realizable moments generated by the particle density and flux vector is also convex, and resembles a ``pinched'' cone.  
It coincides with the realizable set for the classical case when the occupancy is low, but is much more restricted for high occupancy.

In this paper, the two-moment model is discretized in space using high-order Discontinuous Galerkin (DG) methods (e.g., \cite{cockburnShu_2001,hesthavenWarburton_2008}).  
DG methods combine elements from both spectral and finite volume methods, and are an attractive option for solving hyperbolic partial differential equations (PDEs).  
They achieve high-order accuracy on a compact stencil; i.e., data is only communicated with nearest neighbors, regardless of the formal order of accuracy, which can lead to a high computation to communication ratio, and favorable parallel scalability on heterogeneous architectures has been demonstrated \cite{klockner_etal_2009}.  
Furthermore, they can easily be applied to problems involving curvilinear coordinates (e.g., beneficial in numerical relativity \cite{teukolsky_2016}).  
Importantly, DG methods, exhibit favorable properties when collisions with a background are included as they recover the correct asymptotic behavior in the diffusion limit, characterized by frequent collisions (e.g., \cite{larsenMorel_1989,adams_2001,guermondKanschat_2010}).  
The DG method was introduced in the 1970s by Reed \& Hill \cite{reedHill_1973} to solve the neutron transport equation, and has undergone remarkable developments since then (see, e.g., \cite{shu_2016}, and references therein).

We are concerned with the development and application of DG methods for the fermionic two-moment model that can preserve the aforementioned algebraic constraints and ensure realizable moments, provided the initial condition is realizable.  
Our approach is based on the constraint-preserving (CP) framework introduced in \cite{zhangShu_2010a}, and later extended to the Euler equations of gas dynamics in \cite{zhangShu_2010b}.  
(See e.g., \cite{xing_etal_2010,zhangShu_2011,olbrant_etal_2012,cheng_etal_2013,zhang_etal_2013,endeve_etal_2015,wuTang_2015}, for extensions and applications to other systems.)  
The main ingredients include (1) a realizability-preserving update for the cell averaged moments based on Euler time stepping, which evaluates the polynomial representation of the DG method in a finite number of quadrature points in the local elements and results in a Courant-Friedrichs-Lewy (CFL) condition on the time step; (2) a limiter to modify the polynomial representation to ensure that the algebraic constraints are satisfied point-wise without changing the cell average of the moments; and (3) a time stepping method that can be expressed as a convex combination of Euler steps, and therefore preserves the algebraic constraints as the Euler methods (possibly with a modified CFL condition).

The DG discretization leaves the temporal dimension continuous.  
This semi-discretization leads to a system of ordinary differential equations (ODEs), which can be integrated with standard ODE solvers (i.e., the method of lines approach to solving PDEs).  
We use implicit-explicit (IMEX) Runge-Kutta (RK) methods \cite{ascher_etal_1997,pareschiRusso_2005} to integrate the two-moment model forward in time.  
This approach is motivated by the fact that we can resolve time scales associated with particle streaming terms in the moment equations, which will be integrated with explicit methods, while terms associated with collisional interactions with the background induce fast time scales that we do not wish to resolve, and will be integrated with implicit methods.  
This splitting has some advantages when solving kinetic equations since the collisional interactions may couple across momentum space, but are local in position space, and is easier to parallelize than a fully implicit approach.

The CP framework of \cite{zhangShu_2010a} achieve high order (i.e., greater than first order) accuracy in time by employing explicit strong stability-preserving Runge-Kutta (SSP-RK) methods \cite{shuOsher_1988,gottlieb_etal_2001}, which can be written as a convex combination of forward Euler steps.  
Unfortunately, this strategy to achieve high order temporal accuracy does not work as straightforwardly for standard IMEX Runge-Kutta (IMEX-RK) methods because implicit SSP Runge-Kutta methods with greater than first order accuracy do not exist \cite{gottlieb_etal_2001}.  
To break this ``barrier'', recently proposed IMEX-RK schemes \cite{chertock_etal_2015,hu_etal_2018} have resorted to first order accuracy in favor of the SSP property in the standard IMEX-RK scheme, and recover second-order accuracy with a correction step.

We consider the application of the correction approach to the two-moment model.  
However, with the correction step from \cite{chertock_etal_2015} we are unable to prove the realizability-preserving property without invoking a too restrictive time step.  
With the correction step from \cite{hu_etal_2018} the realizability-preserving property is guaranteed with a time step comparable to that of the forward Euler method applied to the explicit part of the scheme, but the resulting scheme performs poorly in the asymptotic diffusion limit.  
Because of these challenges, we resort to first-order temporal accuracy, and propose IMEX-RK schemes that are constraint-preserving with a time step equal to that of forward Euler on the explicit part, perform well in the diffusion limit, and reduce to a second-order SSP-RK scheme in the streaming limit (no collisions with the background material).

The realizability-preserving property of the DG-IMEX scheme depends sensitively on the adopted closure procedure.  
Employing the simple Lax-Friedrichs flux and imposing a suitable CFL condition on the time step, the explicit update of the cell average can be written as a convex combination.  
Realizability of the updated cell average is then guaranteed from convexity arguments \cite{zhangShu_2010a}, provided all the elements in the convex combination are realizable.  
Realizability of individual elements in the convex combination is conditional on the closure procedure (components of the Eddington tensor must be computed to evaluate numerical fluxes).  
We prove that each element in the convex combination is realizable, provided the moments involved in expressing the elements are moments of a distribution function satisfying the bounds implied by Fermi-Dirac statistics (i.e., $0\le f \le 1$).  
For simple algebraic two-moment closures, which we consider, the so-called Eddington factor is given by an algebraic expression depending on the evolved moments and completely determines the components of the Eddington tensor.  
Realizable components of the Eddington tensor demands that the Eddington factor satisfies strict lower and upper bounds (e.g., \cite{levermore_1984,lareckiBanach_2011}).  
We discuss simple algebraic closures derived from Fermi-Dirac statistics that satisfy these bounds, and demonstrate with numerical experiments that the DG-IMEX scheme preserves realizability of the moments when these closures are used.  
We also demonstrate that further approximations to algebraic two-moment closures for modeling particle systems governed by Fermi-Dirac statistics may give results that are incompatible with a bounded distribution and therefore unphysical.  
The example we consider is the Minerbo closure \cite{minerbo_1978}, which can be obtained as the low occupancy limit of the maximum entropy closure of Cernohorsky \& Bludman \cite{cernohorskyBludman_1994}.

>>>>

In this paper we design numerical methods to solve the two-moment model governing transport of particles obeying Fermi-Dirac statistics (e.g., neutrinos), with the ultimate target being nuclear astrophysics applications (e.g., neutrino transport in core-collapse supernovae and compact binary mergers).  
The numerical method is based on the discontinuous Galerkin (DG) method for spatial discretization and implicit-explicit (IMEX) methods for time integration, and is designed to preserve certain physical constraints of the underlying model.  
The latter property is achieved by considering the spatial and temporal discretization together with the closure procedure for the two-moment model.

In many applications, the particle mean free path is comparable to or exceeds other characteristic length scales in the system under consideration (i.e., the Knudsen number exceeds unity), and non-equilibrium effects may become important.  
In these situations, a kinetic description may be required, where a particle distribution function $f(\vect{p},\vect{x},t)$ is solved for.  
The distribution function, a phase space density depending on momentum $\vect{p}\in\bbR^{3}$ and position $\vect{x}\in\bbR^{3}$, gives at time $t\in\bbR^{+}$ the number of particles in the phase space volume element $d\vect{p}\,d\vect{x}$ (i.e., $dN=f\,d\vect{p}\,d\vect{x}$).  
The evolution of the distribution function is governed by the Boltzmann equation, which states a balance between phase space advection and collisions (see, e.g., \cite{braginskii_1965,chapmanCowling_1970,lifshitzPitaevskii_1981}).

@@

Here we consider a two-moment model for particles governed by Fermi-Dirac statistics.  
It is well known from the two-moment model for particles governed by Maxwell-Boltzmann statistics (``classical'' particles with $f\ge0$), that the particle density is nonnegative and the magnitude of the flux vector is bounded by the particle density.  
(There are further constraints on the components of the Eddington tensor \cite{levermore_1984}.)  
Furthermore, the set of realizable moments generated by the particle density and flux vector constitutes a convex cone \cite{olbrant_etal_2012}.  
In the fermionic case, there is also an upper bound on the distribution function (e.g., $f\le1$) because Pauli's exclusion principle prevents particles from occupying the same microscopic state.  
The fermionic two-moment model has recently been studied theoretically in the context of maximum entropy closures \cite{lareckiBanach_2011,banachLarecki_2013,banachLarecki_2017b} and Kershaw-type closures \cite{banachLarecki_2017a}.  
Because of the upper bound on the distribution function, the algebraic constraints on realizable moments differ from the classical case with no upper bound, and can lead to significantly different dynamics when the occupancy is high (i.e., when $f$ is close to its upper bound).  
In the fermionic case, the set of realizable moments generated by the particle density and flux vector is also convex, and resembles a ``pinched'' cone (as will be shown; cf. Figure~\ref{fig:RealizableSetFermionic} in Section~\ref{sec:realizability}).  
It coincides with the realizable set for the classical case when the occupancy is low, but is much more restricted for high occupancy.

In this paper, the two-moment model is discretized in space using high-order Discontinuous Galerkin (DG) methods (e.g., \cite{cockburnShu_2001,hesthavenWarburton_2008}).  
DG methods combine elements from both spectral and finite volume methods, and are an attractive option for solving hyperbolic partial differential equations (PDEs).  
They achieve high-order accuracy on a compact stencil; i.e., data is only communicated with nearest neighbors, regardless of the formal order of accuracy, which can lead to a high computation to communication ratio, and favorable parallel scalability on heterogeneous architectures has been demonstrated \cite{klockner_etal_2009}.  
Furthermore, they can easily be applied to problems involving curvilinear coordinates (e.g., beneficial in numerical relativity \cite{teukolsky_2016}).  
Importantly, DG methods exhibit favorable properties when collisions with a background are included, as they recover the correct asymptotic behavior in the diffusion limit, characterized by frequent collisions (e.g., \cite{larsenMorel_1989,adams_2001,guermondKanschat_2010}).  
The DG method was introduced in the 1970s by Reed \& Hill \cite{reedHill_1973} to solve the neutron transport equation, and has undergone remarkable developments since then (see, e.g., \cite{shu_2016} and references therein).

We are concerned with the development and application of DG methods for the fermionic two-moment model that can preserve the aforementioned algebraic constraints and ensure realizable moments, provided the initial condition is realizable.  
Our approach is based on the constraint-preserving (CP) framework introduced in \cite{zhangShu_2010a}, and later extended to the Euler equations of gas dynamics in \cite{zhangShu_2010b}.  
(See, e.g., \cite{xing_etal_2010,zhangShu_2011,olbrant_etal_2012,cheng_etal_2013,zhang_etal_2013,endeve_etal_2015,wuTang_2015} for extensions and applications to other systems.)  
The main ingredients include (1) a realizability-preserving update for the cell averaged moments based on Euler time stepping, which evaluates the polynomial representation of the DG method in a finite number of quadrature points in the local elements and results in a Courant-Friedrichs-Lewy (CFL) condition on the time step; (2) a limiter to modify the polynomial representation to ensure that the algebraic constraints are satisfied point-wise without changing the cell average of the moments; and (3) a time stepping method that can be expressed as a convex combination of Euler steps, and therefore preserves the algebraic constraints as do the Euler methods (possibly with a modified CFL condition).

The DG discretization leaves the temporal dimension continuous.  
This semi-discretization leads to a system of ordinary differential equations (ODEs), which can be integrated with standard ODE solvers (i.e., the method of lines approach to solving PDEs).  
We use implicit-explicit (IMEX) Runge-Kutta (RK) methods \cite{ascher_etal_1997,pareschiRusso_2005} to integrate the two-moment model forward in time.  
This approach is motivated by the fact that we can resolve time scales associated with particle streaming terms in the moment equations, which will be integrated with explicit methods, while terms associated with collisional interactions with the background induce fast time scales that we do not wish to resolve, and will be integrated with implicit methods.  
This splitting has some advantages when solving kinetic equations since the collisional interactions may couple across momentum space, but are local in position space, and are easier to parallelize than a fully implicit approach.

The CP framework of \cite{zhangShu_2010a} achieves high-order (i.e., greater than first order) accuracy in time by employing strong stability-preserving explicit Runge-Kutta (SSP-RK) methods \cite{shuOsher_1988,gottlieb_etal_2001}, which can be written as a convex combination of forward Euler steps.  
Unfortunately, this strategy to achieve high-order temporal accuracy does not work as straightforwardly for standard IMEX Runge-Kutta (IMEX-RK) methods because implicit SSP Runge-Kutta methods with greater than first order accuracy do not exist \cite{gottlieb_etal_2001}.  
To break this ``barrier,'' recently proposed IMEX-RK schemes \cite{chertock_etal_2015,hu_etal_2018} have resorted to first order accuracy in favor of the SSP property in the standard IMEX-RK scheme, and recover second-order accuracy with a correction step.

We consider the application of the correction approach to the two-moment model.  
However, with the correction step from \cite{chertock_etal_2015} we are unable to prove the realizability-preserving property without invoking a too-restrictive time step.  
With the correction step from \cite{hu_etal_2018} the realizability-preserving property is guaranteed with a time step comparable to that of the forward Euler method applied to the explicit part of the scheme, but the resulting scheme performs poorly in the asymptotic diffusion limit.  
Because of these challenges, we resort to first-order temporal accuracy, and propose IMEX-RK schemes that are constraint-preserving with a time step equal to that of forward Euler on the explicit part, perform well in the diffusion limit, and reduce to a second-order SSP-RK scheme in the streaming limit (no collisions with the background material).

The realizability-preserving property of the DG-IMEX scheme depends sensitively on the adopted closure procedure.  
Employing the simple Lax-Friedrichs flux and imposing a suitable CFL condition on the time step, the explicit update of the cell average can be written as a convex combination.  
Realizability of the updated cell average is then guaranteed from convexity arguments \cite{zhangShu_2010a}, provided all the elements in the convex combination are realizable.  
Realizability of individual elements in the convex combination is conditional on the closure procedure (components of the Eddington tensor must be computed to evaluate numerical fluxes).  
We prove that each element in the convex combination is realizable provided the moments involved in expressing the elements are moments of a distribution function satisfying the bounds implied by Fermi-Dirac statistics (i.e., $0\le f \le 1$).  
For algebraic two-moment closures, which we consider, the so-called Eddington factor is given by an algebraic expression depending on the evolved moments and completely determines the components of the Eddington tensor.  
Realizable components of the Eddington tensor demand that the Eddington factor satisfies strict lower and upper bounds (e.g., \cite{levermore_1984,lareckiBanach_2011}).  
We discuss algebraic closures derived from Fermi-Dirac statistics that satisfy these bounds, and demonstrate with numerical experiments that the DG-IMEX scheme preserves realizability of the moments when these closures are used.  
We also demonstrate that further approximations to algebraic two-moment closures for modeling particle systems governed by Fermi-Dirac statistics may give results that are incompatible with a bounded distribution and, therefore, unphysical.  
The example we consider is the Minerbo closure \cite{minerbo_1978}, which can be obtained as the low occupancy limit of the maximum entropy closure of Cernohorsky \& Bludman \cite{cernohorskyBludman_1994}.
